Competing Bandits: The Perils of Exploration under Competition

### 2018-09-01 paper/content/inverted_u.tex
restructure the paper a bit

\section{Competition vs Better Algorithms: Inverted-U}\label{section:5}

In this section we document our main results that can be summarized via the empirically documented inverted-U relationship between competition and innovation described in \cite{Aghion-QJE05}. We consider three separate settings for our model: permanent monopoly, temporary monopoly (incumbent vs entrant), and permanent duopoly.

\begin{finding}
\textit{Under the priors that are relative reputation costly and with low warm start, the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}

We interpret the move from permanent monopoly to temporary monopoly and temporary monopoly to permanent duopoly as increases in competition. We believe this is reasonable and consistent with the empirical results which define increases in competition as increases in a measure of market power (1 - the Lerner Index). In our model we have no prices and no costs and so we interpret increases in "market power" as decreases in the number of rounds where a firm has to compete for agents.

We utilize the same instances and realizations as in the section 4 and simulate the model described in section 2. We initially take the strategies of the firms as exogenous (chosen from $\mathcal{A}$) and simulate the model given these strategies in order to determine the expected payoffs associated with each pair of algorithms. Unless otherwise noted, all the results are reported at $t = 2000$.

\textbf{Permanent Monopoly} Under permanent monopoly only a single firm exists in the market for every period. Since, in our formulation, the firm only gets utility from a larger market share it is indifferent between each algorithm in $\mathcal{A}$ since, regardless of the algorithm it deploys, it will get the entire market. If we suppose that deploying the "better" algorithms has even an $\epsilon > 0$ cost, then the firm would choose to deploy $DG$.

\textbf{Permanent Duopoly} Permanent duopoly corresponds to the model described in section 2 where $X = 0$ so that both firms enter the market simultaneously. What algorithms are firms incentivized to deploy? We report the average market share taken over the $N$ simulations for a set of exogenous values of $k$. Additionally, we report the mean and median of an additional quantity which we define as the effective end of game.

@@

\begin{finding}
\textit{For low warm starts, when exploration is relative reputation costly $DG >TS$ in competition, whereas when exploration is not relative reputation costly $\TS > DG$. For sufficiently large warm starts, $\TS$ is dominant across all priors.}
\end{finding}

Table \ref{sim_unif} shows the results under the Uniform prior. For the low values of the warm start we can see that $DG > TS$, but that for larger warm starts, $\TS > DG$ \footnote{The same results hold for the Heavy Tail prior and are reported in the supplemental material}. Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $k$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $k$. The intuition for the result in competition aligns with the fact that, for Uniform and Heavy Tail, $\TS$ is relative reputation costly compared to $DG$ and so, for low warm starts, $\TS$ has not yet recovered from the relative reputation loss due to purposeful exploration and so $DG$ beats $\TS$. However, if we move the warm start sufficiently high so that $\TS$ moves to the relative reputation benefit regime, $\TS > DG$.

However, we do not always see that $DG > TS$ under some warm starts. Table \ref{sim_nih} shows the results for the Needle In Haystack prior where we see that $\TS > DG$. This is precisely due to the fact that, under the Needle In Haystack, $\TS$ is not relative reputation costly compared to $DG$ and thus we see that $\TS > DG$.

Combining these results with the fact that the EEOG values are relatively low across all warm starts \footnote{Interestingly, the EEOG also appears to be skewed to the right, similar to the distribution of reputation differences}, these observations seem to imply that understanding the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotically.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table}[ht]
\centering
\caption{Duopoly Experiment Uniform}
\begin{tabular}{rlll}
  \hline
 & k = 20 & k = 250 & k = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 230\\ med: 0} & \makecell{\textbf{0.52} $\pm$0.02\\ eeog \\ avg: 800\\ med: 754} & \makecell{\textbf{0.6} $\pm$0.02\\ eeog \\ avg: 910\\ med: 906.5} \\
  TS vs DEG & \makecell{\textbf{0.41} $\pm$0.03\\ eeog \\ avg: 180\\ med: 0} & \makecell{\textbf{0.51} $\pm$0.02\\ eeog \\ avg: 810\\ med: 734} & \makecell{\textbf{0.55} $\pm$0.02\\ eeog \\ avg: 970\\ med: 987} \\
  DG vs DEG & \makecell{\textbf{0.51} $\pm$0.03\\ eeog \\ avg: 470\\ med: 57.5} & \makecell{\textbf{0.48} $\pm$0.02\\ eeog \\ avg: 1000\\ med: 1088} & \makecell{\textbf{0.45} $\pm$0.02\\ eeog \\ avg: 1000\\ med: 1142} \\
   \hline
\end{tabular}
\label{sim_unif}
\end{table}

\begin{table}[ht]
\centering
\caption{Duopoly Experiment Needle In Haystack}
\begin{tabular}{rlll}
  \hline
 & k = 20 & k = 250 & k = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{\tiny{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95 \% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\textbf{Temporary Monopoly} We now consider asymmetries in the timing of entry so that one firm enters the market before the other and serves as a monopolist in the periods until the other firm enters. In terms of our model, this corresponds to varying the value of $X$. For this section we report results fixing $k = 20$ and focus on the priors where we observed that, for this warm start, $DG > TS$ \footnote{However, we report the results for the Needle In Haystack prior in the supplementary material. The results are unsurprising, where $\TS$ is the dominant strategy for both the entrant and incumbent.}. It seems clear that the temporary monopoly will get a larger market share regardless of what algorithm it plays, but is the incumbent incentivized to commit to $\TS$?

@@

In table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $DG$ is the dominant strategy for the entrant on the Heavy Tail prior \footnote{The result for the Uniform is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$. However, while $DG > TS$ for the entrant, $DG$ does roughly the same as $\DEG$ on the Uniform prior. The results are in the supplementary material.}. In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across priors.

@@

Why do we observe that $\TS$ is the dominant strategy for the incumbent whereas in the permanent duopoly experiment we saw that, under the same other conditions, $DG$ was preferred to $\TS$? The intuition for this is that competition in the duopoly forces the firms to worry about their reputation which dissuades them from committing to algorithms that involve pure exploration in the early rounds. This intuition is very similar to that observed in the previous section by increasing warm start. One can view allowing one firm to temporarily be a monopolist as temporarily relaxing the "incentive" component of exploration, exploitation, and incentives so that the incumbent firm only faces the classic tradeoff between exploration and exploitation. The incumbent only needs to worry about her reputation after $X$ periods when the entrant comes into the market and again needs to worry about incentivizing agents to select them over their competition. As a result, the incumbent is incentivized to commit to an algorithm that does exploration in the early rounds since she no longer suffers the same relative reputational cost that she would suffer under competition as long as the $X$ is sufficiently large that she can begin to recover the reputational costs of exploration. Thus, counterintuitively, by having one firm be a monopoly and dominate the market, we can incentivize them to play $\TS$.

>>>>

\section{Competition vs Better Algorithms}\label{section:5}

In this section we document the results of the competition game for various parameterizations of our model. We consider three separate settings for our model: permanent monopoly, temporary monopoly (incumbent vs entrant), and permanent duopoly.

\OMIT{main results that can be summarized via the empirically documented inverted-U relationship between competition and innovation described in \cite{Aghion-QJE05}. 
}

\OMIT{
\begin{finding}
\textit{For sufficiently low warm start, the optimal strategies in the competition game are:
\begin{center}
\textbf{Permanent Monopoly} - $\DG$ \\
\textbf{Temporary Monopoly} - $\TS$ \\
\textbf{Permanent Duopoly} - $\DG$ \\
\end{center}
The conditions on $\TS$ being dominant under the temporary monopoly are that the incumbent is a temporary monopoly for sufficiently many periods.
}
\end{finding}
}

We interpret the move from permanent monopoly to temporary monopoly and temporary monopoly to permanent duopoly as increases in competition. We believe this is reasonable and consistent with the empirical results which define increases in competition as increases in a measure of market power (1 - the Lerner Index). In our model we have no prices and no costs and so we interpret increases in ``market power" as decreases in the number of rounds where a firm has to compete for agents.

% couldn't get section references to work so hardcoding for now
We utilize the same instances and realizations as in the section 3  and simulate the model described in section 2. We initially take the strategies of the firms as exogenous (chosen from $\mathcal{A}$) and simulate the model given these strategies in order to determine the expected payoffs associated with each pair of algorithms. Unless otherwise noted, all the results are reported at $t = 2000$.

\textbf{Permanent Monopoly} Under permanent monopoly only a single firm exists in the market for every period. In our model the firm only gets utility from a larger market share so it is indifferent between each algorithm in $\mathcal{A}$ since, regardless of the algorithm it deploys, it will get the entire market. If we suppose that deploying the ``better" algorithms has even an $\epsilon > 0$ cost or, equivalently, that firms break indifference towards algorithms that are easier to deploy, then the firm would choose to deploy $\DG$.

\textbf{Permanent Duopoly} Permanent duopoly corresponds to the model described in section 2 where $X = 0$ so that both firms enter the market simultaneously. What algorithms are firms incentivized to deploy? We report the average market share taken over the $N$ simulations for a set of exogenous values of $T_0$. Additionally, we report the mean and median of an additional quantity which we define as the \textit{effective end of game}.

@@

\begin{finding}
\textit{
\begin{enumerate}
\item For low warm starts, under Heavy Tail and Uniform \\ $\DG > \TS$ in competition (see Table \ref{sim_ht}),
\item Under Needle In Haystack, \\ $\TS > \DG$ (see Table \ref{sim_nih}). 
\item For sufficiently large warm starts, $\TS$ is dominant across all the families of instances.
\item Even for a large warm start, $\DG > \DEG$ under Heavy Tail
\end{enumerate}
}
\end{finding}

\OMIT{
Looking at the relative reputation plots in Figure \ref{relative_rep_plots}, we can interpret fixing a warm start $T_0$ as fixing the starting point on the relative reputation plots. The proportion of first rounds in the competition game that will go to a firm playing alg $A$ over a firm playing alg $B$ will correspond to the relative reputation proportion at time $k$.
The intuition for the result in competition aligns with the fact that, for Uniform and Heavy Tail, $\TS$ is relative reputation costly compared to $\DG$ and so, for low warm starts, $\TS$ has not yet recovered from the relative reputation loss due to purposeful exploration and so $\DG$ beats $\TS$. However, if we move the warm start sufficiently high so that $\TS$ moves to the relative reputation benefit regime, $\TS > \DG$.
}

 \OMIT{However, we do not always see that $\DG > \TS$ under some warm starts. Table \ref{sim_nih} shows the results for the Needle In Haystack family where we see that $\TS > \DG$. This is precisely due to the fact that, under the Needle In Haystack, $\TS$ is not relative reputation costly compared to $\DG$ and thus we see that $\TS > \DG$.}

Combining this finding with the fact that the EEOG values are relatively low across all warm starts \footnote{Interestingly, the EEOG also appears to be skewed to the right}, these observations seem to imply that understanding the performance of different algorithms in competition it is important to look at their performance for relatively small samples instead of asymptotically.

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\begin{table}[ht]
\centering
\caption{Duopoly Experiment Heavy Tail} 
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\ 
  \hline
TS vs DG & \makecell{\textbf{0.29} $\pm$0.03\\ eeog \\ avg: 55\\ med: 0} & \makecell{\textbf{0.72} $\pm$0.02\\ eeog \\ avg: 570\\ med: 0} & \makecell{\textbf{0.76} $\pm$0.02\\ eeog \\ avg: 620\\ med: 98.5} \\ 
  TS vs DEG & \makecell{\textbf{0.3} $\pm$0.03\\ eeog \\ avg: 37\\ med: 0} & \makecell{\textbf{0.88} $\pm$0.01\\ eeog \\ avg: 480\\ med: 0} & \makecell{\textbf{0.9} $\pm$0.01\\ eeog \\ avg: 570\\ med: 113.5} \\ 
  DG vs DEG & \makecell{\textbf{0.62} $\pm$0.03\\ eeog \\ avg: 410\\ med: 7} & \makecell{\textbf{0.6} $\pm$0.02\\ eeog \\ avg: 790\\ med: 762} & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 730\\ med: 608} \\ 
   \hline
\end{tabular}
\label{sim_ht}
\end{table}

\begin{table}[ht]
\centering
\caption{Duopoly Experiment Needle In Haystack}
\begin{tabular}{rlll}
  \hline
 & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
TS vs DG & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 200\\ med: 27} & \makecell{\textbf{0.6} $\pm$0.03\\ eeog \\ avg: 370\\ med: 0} & \makecell{\textbf{0.64} $\pm$0.03\\ eeog \\ avg: 580\\ med: 121.5} \\
  TS vs DEG & \makecell{\textbf{0.57} $\pm$0.03\\ eeog \\ avg: 150\\ med: 14} & \makecell{\textbf{0.52} $\pm$0.03\\ eeog \\ avg: 460\\ med: 78.5} & \makecell{\textbf{0.56} $\pm$0.02\\ eeog \\ avg: 740\\ med: 627.5} \\
  DG vs DEG & \makecell{\textbf{0.46} $\pm$0.03\\ eeog \\ avg: 340\\ med: 128.5} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 650\\ med: 408} & \makecell{\textbf{0.42} $\pm$0.02\\ eeog \\ avg: 690\\ med: 466.5} \\
   \hline
\end{tabular}
\label{sim_nih}
\caption*{\tiny{The first line in each cell contains the average market share received by the firm playing Alg 1 (and the market share of Alg 2 is 1 - Alg 1 Market Share) as well as a 95 \% confidence band. For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG. The next line contain the average and median effective end of game for this set of simulations.}}
\end{table}
% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:00 2018

\textbf{Temporary Monopoly} We now consider asymmetries in the timing of entry so that one firm enters the market before the other and serves as a monopolist in the periods until the other firm enters. In terms of our model, this corresponds to varying the value of $X$. For this section we report results fixing $T_0 = 20$ and focus on the family of instances where we observed that, for this warm start, $\DG > \TS$ \footnote{However, we report the results for the Needle In Haystack family in the supplementary material. The results are unsurprising, where $\TS$ is the dominant strategy for both the entrant and incumbent.}. It seems clear that the temporary monopoly will get a larger market share regardless of what algorithm it plays, but is the incumbent incentivized to commit to $\TS$?

@@

In table \ref{ht_incum} we can see that, for $X = 200$, $\TS$ is a dominant strategy for the incumbent and that $\DG$ is the dominant strategy for the entrant on the Heavy Tail family \footnote{The result for the Uniform is the same, though it takes a higher $X$ to induce the incumbent to play $\TS$. However, while $\DG > \TS$ for the entrant, $\DG$ does roughly the same as $\DEG$ on the Uniform family. The results are in the supplementary material.}. In the supplementary material we report the same experiment for different values of $X$ and show that our main result that, for sufficiently large $X$, $\TS$ is the dominant strategy for the incumbent is robust across the family of instances we consider.

@@

Why do we observe that $\TS$ is the dominant strategy for the incumbent whereas in the permanent duopoly experiment we saw that, under the same parameters, $\DG$ was preferred to $\TS$? The intuition for this is that competition in the duopoly forces the firms to worry about their reputation which dissuades them from committing to algorithms that involve pure exploration in the early rounds. One can view allowing one firm to temporarily be a monopolist as temporarily relaxing the ``incentive" component of exploration, exploitation, and incentives so that the incumbent firm only faces the classic tradeoff between exploration and exploitation. The incumbent only needs to worry about her reputation after $X$ periods when the entrant comes into the market and again needs to worry about incentivizing agents to select them over their competition. \OMIT{As a result, the incumbent is incentivized to commit to an algorithm that does exploration in the early rounds since she no longer suffers the same relative reputational cost that she would suffer under competition as long as the $X$ is sufficiently large that she can begin to recover the reputational costs of exploration.} Thus, counterintuitively, by having one firm be a monopoly and dominate the market, we can incentivize them to play $\TS$.

### 2018-09-01 paper/content/perf_in_iso.tex
restructure the paper a bit

We start with a pilot experiment in which we investigate each algorithm's performance "in isolation": in a stand-alone MAB problem without competition. We focus on reputation scores generated by each algorithm.

First, we confirm that algorithms' performance is ordered as we'd expect:
    $\TS \gg \DEG \gg \DG$.
For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. Figure \ref{prelim_means} shows such plot for the Needle In Haystack instance (for other MAB instances the plots are similar). We summarize this finding as follows:

@@

The mean reputation trajectory is, arguably, the most natural way to represent algorithm's performance on a given problem instance. But is it the right statistic for our purposes? That is, how well do these trajectories predict the outcomes of the competition game? The following conjecture is natural:

\begin{conjecture}\label{conj:mean-trajectories}
For a given MAB instance, if one algorithm's mean reputation trajectory lies above another, then the first algorithm prevails in the competition game.
\end{conjecture}

However, we find a more nuanced picture.

@@

@@

To see what could go wrong, consider how algorithm's reputation score is distributed at a particular time. That is, consider the empirical distribution of this score over different \MRVs. For concreteness, consider the needle-in-haystack instance at time $t=500$ (Figure \ref{rep_dist_nih}). 
We see that the ``naive" algorithms $\DG$ and $\DEG$ have a bi-modal reputation distribution, whereas $\TS$ does not. The reason is that for this MAB instance, $\DG$ either finds the best arm and sticks to it, or gets stuck on the bad arms.
However, the mean reputation trajectory of an algorithm would summarize all this complexity the average over the \MRVs. This may be inadequate for explaining the outcome of the duopoly game, given that the latter is determined by a simple comparison between the firm's reputation scores. Further, consider the difference in reputation scores (\emph{reputation difference}) between \TS and \DG on a particular \MRV. Let's plot the empirical distribution of the reputation difference (over the \MRVs) at a particular time point. Figure~\ref{ts_dg_rep_diff_nih} shows such plots for several time points. We see that the distribution is skewed to the right, for which the mean is not a representative statistic.  \ascomment{so what?}

@@

>>>>

We start with a pilot experiment in which we investigate each algorithm's performance ``in isolation": in a stand-alone MAB problem without competition. We focus on reputation scores generated by each algorithm.

First, we confirm that\gaedit{, for a sufficiently long time horizon,} algorithms' performance is ordered as we'd expect:
    $\TS \gg \DEG \gg \DG$.
For each algorithm and each MAB instance, we compute the mean reputation score at each round, averaged over all \MRVs. We plot the \emph{mean reputation trajectory}: how this score evolves over time. Figure \ref{prelim_means} shows such plot for the Needle In Haystack instance (for other MAB instances the plots are similar). We summarize this finding as follows:

@@

The mean reputation trajectory is, arguably, the most natural way to represent an algorithm's performance on a given problem instance. But is it the right statistic for our purposes? That is, how well do these trajectories predict the outcomes of the competition game? The following conjecture is natural:

\begin{conjecture}\label{conj:mean-trajectories}
For a given MAB instance, if one algorithm's mean reputation trajectory lies above another at time $T_0$, then the first algorithm prevails in the competition game.
\end{conjecture}

We revisit this conjecture after documenting the results in the competition game.

\OMIT{
However, we find a more nuanced picture.

@@

We will revisit this finding after presenting the results from the competition game.

@@

To see what could go wrong, consider how an algorithm's reputation score is distributed at a particular time. That is, consider the empirical distribution of this score over different \MRVs. For concreteness, we consider the needle-in-haystack instance at time $t=500$ (Figure \ref{rep_dist_nih}).
We see that the ``naive" algorithms $\DG$ and $\DEG$ have a bi-modal reputation distribution, whereas $\TS$ does not. The reason is that for this MAB instance, $\DG$ either finds the best arm and sticks to it, or gets stuck on the bad arms. This leads to two cases: $\DG$ either does slightly better than $\TS$ or where $\TS$ does substantially better than $\DG$.

However, the mean reputation trajectory of an algorithm would summarize all this complexity the average over the \MRVs. This may be inadequate for explaining the outcome of the duopoly game, given that the latter is determined by a simple comparison between the firm's reputation scores. Further, consider the difference in reputation scores (\emph{reputation difference}) between \TS and \DG on a particular \MRV. Let's plot the empirical distribution of the reputation difference (over the \MRVs) at a particular time point. Figure~\ref{ts_dg_rep_diff_nih} shows such plots for several time points. We see that the distribution is skewed to the right, for which the mean is not a representative statistic.  \ascomment{so what?}

@@

}
